How to improve your relationship with your future self, V 2.0

### 2016-06-28 workflow.Rmd
Almost done

Jake is less organized about directory structure and tends to have directories with structures like this:

```{bash, tidy=FALSE, eval=FALSE}

manuscript\

@@

Group work requires version control. Many people are familiar with the track changes
feature in modern WYSIWYG (what you see is what you get) word processors or the fact that Dropbox allows one
to recover previous versions of files. These are both kinds of version control.
More generally, when we collaborate, we'd like to do a variety of actions with
our shared files. Collaboration on data analytic projects is more productive
and better when (1) it is easy to see what has changed between versions of
files; (2) members of the team feel free to experiment and then to dump parts
of the experimentation in favor of previous work while merging the successful parts
 into the main body of the paper; (3) the team can produce
have 'releases' of the same document (one to MPSA, one to APSR, one to your
parents) without spawning many possibly conflicting copies of the same
document; (4) people can work on the same files at the same time without
conflicting with one another, and can reconcile their changes without too much
confusion and clobbering. Clobbering is what happens when your future self or
your current collaborator saves an old version of a file over a new version,
erasing good work by accident.

Of course if you rely on Dropbox or 'track changes' in Word for version
control, you must communicate with other folks in your group before you edit
existing files. Only one of you can edit and save a given file at a time. This
prevents your work (or your colleagues work) from getting lost when you both
try to save the same file on top of each other. If you find that you need to
work on the same files at the same time, then you should work on establishing
your own shared version control system. Free options include launchpad, github, Open Science Framework,
sourceforge for open source projects (i.e. papers you are writing which you are
happy to share with others as you write, see more below). Each has a slightly difference language (some say folders other repositories, some say download othere clone, etc) and it takes a bit of time adjusting. But as we'll argue: it is well worth the time investment as it will save (lots!) of time later on. Currently Google Docs does
an excellent job with shared editing of files. The downside is: you have to be online. For people with irregular or slow internet connections, this may not be an option. MORE ON GOOGLE?
One may also use Dropbox as a kind of server for version control (though the downside is if your collaborations or project grows large (in terms of MBs) you need to switch to a paid service). Maarten has been using this for most of his collaborations. You can copy files from the Dropbox
directory into a local working directory so as to avoid clobbering and then
work on merging changes by hand before copying back to the Dropbox directory
and replacing existing files. You will however need to agree on an iron file and directory naming rule with your collaborators.

Jake uses github within his research group, and used it for all of
his own projects. <!-- JAKE do you want to add something else here? -->
You may even obtain a two year unlimited repositories with an academic affiliation.

An excellent, simple, and robust version control system is to rename your files with the date and time of
saving them: yyyymmdd_project.docx (for changes within the same day, just before a deadline for example, you may even add a time, so it becomes 20160623_4PM_inequalitypaper.docx). This communicates very clear what version you are working. Remember the days when you received a file called inequalityMV4_APSRsubmit_reallyfinal2.tex? That should quickly become something of the past! We find ourselves preaching the gospel to our collaborators frequently and will keep renaming files until a collaborator has been successfully pacified. Be sure to include year in the file names --- remember, the life of an idea is
measured in years. If you are wise enough to have saved your documents
as plain text then you can easily compare documents using the many
utilities available for comparing text files.^[Adobe Acrobat
  allows one to compare differences in pdf files. OpenOffice supports
  a 'Compare Documents 'option. Word now does the same. And Google Docs will report on the
  version history of a document.] When you reach certain milestones
you can rename the file accordingly: thedocAPSA2009.tex --- for the
one sent to discussants at APSA --- or thedocAPSR2015.tex --- for the
version eventually sent to the APSR six years after you presented it
at APSA. The formal version control systems mentioned above all
allow this kind of thing and are much more elegant and capable, but
you can do it by hand too as long as you don't mind taking up a lot of
disk space and having many 'thedoc... ' files around. Indeed, the number of files can add up quickly (especially around submission time!) and you may want to create an "0_Archive" folder to store older versions. If you do
version control by hand, spend a little extra time to ensure that you
do not clobber files when you make mistakes typing in the
file-names. And, if you find yourself spending extra time reconciling
changes made by different collaborators by hand, remember this is a
task that modern version control systems take care of quickly and
easily.

**Step 5** Writing is rewriting. Thus, all writing
involves versions. When we collaborate with ourselves and others we
want to avoid clobbering and we want to enable graceful reconciliation
of rewriting. One can do these things with formal systems of software
(like subversion or git or bazaar) or with formal systems of file
naming, file comparing, and communication or, even better, with
both. In either case, plain text files will make such tasks easier,
will take up less disk space, and be easier to read for the future
you.

@@

Anyone writing custom code should worry about getting it right. The more code one writes, the more time one has to appreciate problems arising from bugs, errors, and typos in data analysis and code. Mistakes will always be made we should just recognize we need more standardization of practice to minimize the mistakes and help others find them. One thing we can do is to include testing in our coding.

@@

Of course, this idea, like most others, is not new. The desire to
avoid error looms large when large groups of programmers write code
for multi-million dollar programs. The idea of
[test driven development](http://en.wikipedia.org/wiki/Test-driven_development) and the idea that one ought to create tests of
[small parts of one's code](http://en.wikipedia.org/wiki/Unit_testing) arose to address such concerns.^[There are now R packages to help R package writers do this, see for example <https://github.com/hadley/testthat> and the article on it <https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf>.] For the social scientist collaborating with her future self and/or a small group of
collaborators, here is an example of this idea in a very simple form:
Say you want to write a function to multiply a number by 2. If the
function works, when you give it the number 4, you should see it return
the number 8 and when you give it -4, you should get -8.

```{r unit.test}
##  The test function:
test.times.2.fn <- function(){
  ##  This function tests times.2.fn
  if (times.2.fn(thenumber=4) ==  8 &
      times.2.fn(thenumber=-4) == -8) {
    print("It works!")
  } else { print("It does not work!")
         }
}

##  The function:
times.2.fn <- function(thenumber){
  ##  This function multiplies a scalar number by 2
  ##  thenumber is a scalar number
  thenumber+2
}

##  Use the test function
test.times.2.fn()
```

Ack! we mistyped '+' for '*'. Good thing we wrote the
test!^[A more common example of this kind of testing occur
  everyday when we recode variables into new forms but look at a
  crosstab of the old vs. new variable before proceeding.]

MAYBE SOMETHING HERE ON `stopifnot` and recoding data.

@@

Lots of people are thinking about 'reproducible research', creating a 'reproducible workflow' and
'literate programming' these days. Google those terms. Of course, the
devil is in the details: Here I list a few of my own attempts at enabling
reproducible research. You'll find many other inspiring examples on
the web. Luckily, the open source ethos aligns nicely with academic
incentives, so we are beginning to find more and more people offering
their files online for copying and improvement. By the way, if you do
copy and improve, it is polite to alert the person from whom you made
the copy about your work.

We have experimented with several systems so far: (1) we wrote this paper in R markdown (Jake) and Atom (Maarten) and organized our files and collaboration on Github. The source code can be accessed, downloaded and modified from <https://github.com/jwbowers/workflow>. The program also allowed us to send notes to each other. The nice thing about github is that is enables you to make "releases" of a project which enable you to smooth the reproduction of all of the
products of your research (simulations, data analyses, tables, figures, etc..). Similar features are offered by the Open Science Framework (https://osf.io/), Maarten recently participated in a workshop on research transparency organized by the Berkeley Institute for Transparency in the Social Sciences (www.bitss.org), all course material was accessible through OSF (https://osf.io/qh2nr/). Both platforms, OSF and Github nicely integrate and a key benefit is that they have been created to remain for a long time (unlike perhaps your university drive) <!-- ADD REFERENCE TO YOUR OSF PROJECT -->; (2) for one paper Jake simply included a Sweave document and data files into a compressed archive [@bowers2005dataverse] <!-- PERHAPS SAY ABIT MORE ABOUT WHAT SWEAVE IS IN PRINCIPLE-->; (3) for another more computing intensive paper Jake assembled a set of files that enabled reproduction of results using the
`make` system [@bowers2008dataverse] <!-- AGAIN A FEW WORDS ON WHAT PERHAPS LINK TO A WEBSITE? -->; (4) Jake also tried the "compendium"
approach [@gentleman2005reproducible,@gentleman2007statistical] which embeds an
academic paper within the R package system [@bowers2011dataverse]; The benefit of the compendium approach is that one is not required to have
access to a command line for `make`: the compendium is downloadable from within
R using `install.packages()` and is viewable using the `vignette()` function in
any operating system than runs R.^[Notice that Jake's reproduction archives and/or
instructions for using them are hosted on the
[Dataverse](http://thedata.org/book/learn-about-project), which is another
system designed to enhance academic collaboration across time and space.]  The
idea that one ought to be able to install and run and use an academic paper
just as one installs and uses statistical software packages is very attractive,
and we anticipate that it will become ever easier to turn papers into R packages
as creative and energetic folks turn their attention to the question of
reproducible research.^[See for example <http://r-pkgs.had.co.nz/>]
Finally, (5) Maarten has been using Dropbox for most of his collaborative projects (https://www.dropbox.com/). See comments above. As a tool for making files publicly available Dropbox is less useful. You can of course make replication files available through creating an public folder, but you will still need to refer to this folder on a particular project website (an example is Maarten's project on conflict and football https://www.dropbox.com/sh/wqosq99019mf04d/AAASTgJLkn6lmZJI_1veUK7ca?dl=0)

There are some noteworthy new developments also. People at UC Berkeley have been developing a is a web app, Jupyter Notebook (http://jupyter.org/), that enables you to make and share documents that include text, code, equations and visualizations in one INCLUDE REF TO GITHUB - USING PYTHON. For researchers working across different platforms, there is Docker (https://www.docker.com/).

There are also are great notes programs that help you communicate with collaborators. Remember, while email is fantastic for fast communication, it is not a tool designed for project management. It may work for small projects with few co-authors but when the number of collaborators increase it becomes very difficult to keep track. There are some great online systems for task management, most of which have mobile apps also, examples are Flow (www.getflow.com), Asana (www.asana.com), Wrike (www.wrike.com) <!-- JAKE do you know others? --> . There also are great programs for notes. For example, Simplenote is very basic but allows you to share and edit notes. Others are Evernote (www.evernote.com) and for Windows users there is OneNote.

**Step 6** We all learn by doing. When we create a reproducible workflow and share
reproduction materials we improve both cumulation of knowledge and our
methods for doing social science
[@freese2007replication,king1995replication}.

>>>>

Jake tends to be less organized than Maarten and relies on README.md files to tell his future self what is going on in a given directory, a [version control system](http://github.com) to keep track of versions of files, and a [Makefile](http://kbroman.org/minimal_make/) to keep track of relationships between files.

```{bash, tidy=FALSE, eval=FALSE}
manuscript\
figures\
analysis\
  README.md
data\
build\
libraries\
README.md
Makefile
```

@@

Group work requires version control. Many people are familiar with the 'track
changes' feature in modern WYSIWYG (what you see is what you get) word
processors or the fact that Dropbox allows one to recover previous versions of
files. These are both kinds of version control.  More generally, when we
collaborate, we'd like to do a variety of actions with our shared files.
Collaboration on data analytic projects is more productive and better when (1)
it is easy to see what has changed between versions of files; (2) members of
the team feel free to experiment and then to dump parts of the experimentation
in favor of previous work while merging the successful parts into the main body
of the paper; (3) the team can produce 'releases' of the same document
(one to MPSA, one to APSR, one to their parents) without spawning many possibly
conflicting copies of the same document; (4) people can work on the same files
at the same time without conflicting with one another, and can reconcile their
changes without too much confusion and clobbering. Clobbering is what happens
when your future self or your current collaborator saves an old version of a
file over a new version, erasing good work by accident.

Many rely on Dropbox, 'track changes' in Word, or Google Docs for
collaboration. Those tools tend to require you to communicate with other folks
in your group before you edit existing files. On Dropbox, only one of you can
edit and save a given file at a time. Tracking changes with Word tends to
function best if two people are not editing the same document (merging the
changes between two documents is not simple or builtin). Google Docs allows for
two people to edit the same document at the same time: but you must be online
with a fairly fast internet connection.^[Note to future selves: change this
section every five years or so. Stop saying 'internet' in 2026.] Communication
and trading off editing a file prevents your work (or your colleagues' work)
from getting lost when you both try to save the same file on top of each other. One may also use Dropbox as a kind of server for version
control (though the downside is if your collaborations or project grows large
(in terms of MBs) you need to switch to Dropbox's paid service). Maarten has been using
this for most of his collaborations. You can copy files from the Dropbox
directory into a local working directory so as to avoid clobbering and then
work on merging changes by hand before copying back to the Dropbox directory
and replacing existing files. You will however need to agree on an iron file
and directory naming rule with your collaborators to ensure that the files and directories are always named the same across machines and versions.

Another approach to organizing work on many files, including files for code and
for writing, is to use a formal [version
control](https://en.wikipedia.org/wiki/Version_control) strategy. As of 2016,
the standard for managing large collaborations with many files is the
[Git](https://en.wikipedia.org/wiki/Git_(software)) system. User friendly free
front-ends for Git currently make the process of learning and using Git very
convenient and integrated into the kind of workflow that your future self will
be proud of: [Open Science Framework](http://osf.io)
[bitbucket](http://bitbucket.com), [github](http://github.com). Jake uses
[github](http://github.com) for all of his collaborations with others as well
as with his future self: his collaborators appreciate some of the nice extra
features of github, such as the ability to keep a [shared task
list](https://guides.github.com/features/issues/). Learning about version
control systems takes a bit of time. We suggest, however, it is well worth the
time investment as it will save (lots!) of time later on.

An excellent, simple, and robust version control system is to rename your files
with the date and time of saving them: yyyymmdd_project.docx (for changes
within the same day, just before a deadline for example, you may even add a
time, so it becomes 20160623_4PM_inequalitypaper.docx). This communicates very
clear what version you are working. Remember the days when you received a file
called inequalityMV4_APSRsubmit_reallyfinal2.tex? That should quickly become
something of the past! We find ourselves preaching the gospel to our
collaborators frequently and will keep renaming files until a collaborator has
been successfully pacified. Be sure to include year in the file names ---
remember, the life of an idea is measured in years. If you are wise enough to
have saved your documents as plain text then you can easily compare documents
using the many utilities available for comparing text files.^[Adobe Acrobat
allows one to compare differences in pdf files. OpenOffice supports a 'Compare
Documents' option. Word now does the same. And Google Docs will report on the
version history of a document.] When you reach certain milestones you can
rename the file accordingly: thedocAPSA2009.tex --- for the one sent to
discussants at APSA --- or thedocAPSR2015.tex --- for the version eventually
sent to the APSR six years after you presented it at APSA. The formal version
control systems mentioned above all allow this kind of thing and are much more
elegant and capable, but you can do it by hand too as long as you don't mind
taking up a lot of disk space and having many 'thedoc... ' files around.
Indeed, the number of files can add up quickly (especially around submission
time!) and you may want to create an "0_Archive" folder to store older
versions. If you do version control by hand, spend a little extra time to
ensure that you do not clobber files when you make mistakes typing in the
file-names. And, if you find yourself spending extra time reconciling changes
made by different collaborators by hand, remember this is a task that modern
version control systems take care of quickly and easily.

**Step 5** Writing is rewriting. Thus, all writing involves versions. When we
collaborate with ourselves and others we want to avoid clobbering and we want
to enable graceful reconciliation of rewriting. One can do these things with
formal systems of software (like git) or with formal systems of file naming,
file comparing, and communication or, even better, with both. In either case,
plain text files will make such tasks easier, will take up less disk space, and
be easier to read for the future you.

@@

Anyone writing custom code should worry about getting it right. The more code
one writes, the more time one has to appreciate problems arising from bugs,
errors, and typos in data analysis and code. Since mistakes will always be
made, we should just recognize we need more standardization of practice to
minimize the mistakes and help others find them. One thing we can do to
minimize and catch the inevitable mistakes is to include testing in our coding.

@@

Of course, this idea, is not new. The desire to avoid error
looms large when large groups of programmers write code for multi-million
dollar programs. The idea of [test driven
development](http://en.wikipedia.org/wiki/Test-driven_development) and the idea
that one ought to create tests of [small parts of one's
code](http://en.wikipedia.org/wiki/Unit_testing) arose to address such
concerns.^[There are now R packages to help R package writers do this, see for
example <https://github.com/hadley/testthat> and the article on it
<https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf>.]

For the social scientist collaborating with her future self and/or a small
group of collaborators, here is an example of this idea in a very simple form:
Say you want to write a function to multiply a number by 2. If the function
works, when you give it the number 4, you should see it return the number 8 and
when you give it -4, you should get -8.

```{r unit.test, tidy=FALSE}
##  The test function:
test.times.2.fn <- function(){
	##  This function tests times.2.fn
	if (times.2.fn(thenumber=4) ==  8 &
	    times.2.fn(thenumber=-4) == -8) {
		print("It works!")
		} else { print("It does not work!")
	}
}

##  The actual function is:
times.2.fn <- function(thenumber){
	##  This function multiplies a scalar number by 2
	##  thenumber is a scalar number
	thenumber+2
}
```
Here we use the test function to make sure it works.

```{r results='asis'}
test.times.2.fn()
```

Ack! we mistyped '+' for '*'. Good thing we wrote the
test!^[A more common example of this kind of testing occurs
everyday when we recode variables into new forms but look at a
crosstab of the old vs. new variable before proceeding.]

That approach works well when one is paying close attention to messages within
the code. Another approach that we like better is to make the code stop with an
error whenever it doesn't pass a test. In R we use the `stopifnot` function for
this. For example, say we wanted to re-run our analyses excluding the countries
with extreme values of protest. The following code makes a new dataset
excluding places where more than 90% of the respondents to the World Values
survey reported some protest activity. If all of the values of `protest05` are
not less than 90, the code will stop with an error.

```{r}
smalldat <- subset(good.df,subset=protest05<90)
stopifnot(all(smalldat$protest05<90))
```

@@

Lots of people are thinking about 'reproducible research', creating a
'reproducible workflow' and 'literate programming' these days. Google those
terms. Of course, the devil is in the details: Here we list a few of our own
attempts at enabling reproducible research. You'll find many other inspiring
examples on the web. Luckily, the open source ethos aligns nicely with academic
incentives, so we are beginning to find more and more people offering their
files online for copying and improvement. By the way, if you do copy and
improve, it is polite to alert the person from whom you made the copy about
your work and to credit them in some way.

We have experimented with several systems so far:

(1) We wrote this paper in the [R markdown](http://rmarkdown.rstudio.com/)
literate programming format using two different text editors (Jake used vim)
and (Maarten used [atom](https://atom.io/)). We organized our files and
collaboration on Github. The source code for this document can be accessed,
downloaded and modified from <https://github.com/jwbowers/workflow>. We also
used Github Issues to send notes to each other and maintain a task list. The
nice thing about github is that is enables you to make "releases" of a project
which enable you to smooth the reproduction of all of the products of your
research (simulations, data analyses, tables, figures, etc..).  Similar
features are offered by the [Open Science Framework (OSF)](https://osf.io/).
Maarten recently participated in a workshop on research transparency organized
by the [Berkeley Institute for Transparency in the Social
Sciences](www.bitss.org) in which all [course material](https://osf.io/qh2nr/)
was accessible through OSF.  OSF integrates with Github nicely and a key
benefit is that both systems have been created to remain for a long time
(unlike perhaps your university drive).  (2) For one paper Jake simply mixed R
and \LaTeX code in one document (called the "Sweave" format of literate
programming) and then added that document and data into into a compressed
archive [@bowers2005dataverse]; (3) For another more computing intensive paper
Jake assembled a set of files that enabled reproduction of results using the
`make` system [@bowers2008dataverse]; (4) Jake also tried the "compendium"
approach [@gentleman2005reproducible,@gentleman2007statistical] which embeds an
academic paper within the R package system [@bowers2011dataverse]; The benefit
of the compendium approach is that one is not required to have access to a
command line for `make`: the compendium is downloadable from within R using
`install.packages()` and is viewable using the `vignette()` function in any
operating system than runs R.^[Notice that Jake's reproduction archives and/or
instructions for using them are hosted on the
[Dataverse](http://thedata.org/book/learn-about-project), which is another
system designed to enhance academic collaboration across time and space.]  The
idea that one ought to be able to install and run and use an academic paper
just as one installs and uses statistical software packages is very attractive,
and we anticipate that it will become ever easier to turn papers into R
packages as creative and energetic folks turn their attention to the question
of reproducible research.^[See for example <http://r-pkgs.had.co.nz/>] Finally,
(5) Maarten has been using Dropbox for most of his collaborative projects. See
comments above. As a tool for making files publicly available Dropbox is less
useful. You can of course make replication files available through creating an
public folder, but you will still need to refer to this folder on a particular
project website (an example is [Maarten's project on conflict and football](
https://www.dropbox.com/sh/wqosq99019mf04d/AAASTgJLkn6lmZJI_1veUK7ca?dl=0)).

There are some noteworthy new developments also. People at UC Berkeley have
been developing a web app, Jupyter Notebook (http://jupyter.org/), that enables
you to make and share documents that include text, code, equations and
visualizations in one [literate programming
environment](http://blog.jupyter.org/2015/05/07/rendering-notebooks-on-github/).
For researchers working across different platforms, there is
[Docker](https://www.docker.com/) which promises to enable all of the
collaborators on one project to use the same computing environment even if they
are using different laptops running different operating systems.

There are also are great notes programs that help you communicate with collaborators. Remember, while email is fantastic for fast communication, it is not a tool designed for project management. It may work for small projects with few co-authors but when the number of collaborators increase it becomes very difficult to keep track. There are some great online systems for task management, most of which have mobile apps also, examples include [Flow](http://www.getflow.com), [Asana](http://www.asana.com), [Wrike](http://www.wrike.com), [Basecamp](http://basecamp.com) and many others. There also are great programs for notes. For example, Simplenote is very basic but allows you to share and edit notes. Others include [Evernote](www.evernote.com) and for Windows users there is OneNote.

**Step 6** We all learn by doing. When we create a reproducible workflow and share
reproduction materials we improve both cumulation of knowledge and our
methods for doing social science
[@freese2007replication,king1995replication].
